Natural variation in teosinte at the domestication locus teosinte branched1 (tb1)

### 2014-05-01 manuscript.tex
edits

@@

\jri{Check citation format for AJB. I don't think it's numbers. adjust bibliography style accordingly }

@@

\jri{MBH todo} Variation at \emph{tb1} has also been shown to contribute to phenotypes other than tillering \citep{Clark2006}, and a recent study by \citet{StuderDoebley2012} examined the possibility of an allelic series at the \emph{tb1} locus in teosinte. \citet{StuderDoebley2012} introgressed 9 separate teosinte segments (one from \emph{Zea diploperennis}, and four from both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis}) spanning the \emph{tb1} locus into an isogenic maize background and investigated their effects on previously associated phenotypes. They found that plants with teosinte chromosomal segments had greater tillering than their maize isogenic lines, and that different chromosomal segments of \emph{tb1} confer different amounts of tillering, suggesting that there are multiple genetic factors in this region that affect tillering. However, in addition to elucidating variance in tillering among \emph{tb1} teosinte segments, \citet{StuderDoebley2012} found significant variance among W22 control lines, suggesting that there are other genetic factors aside from alleles at the \emph{tb1} locus that affect tillering in maize. \citet{DoebleyStec1991, DoebleyStec1993} first attempted to map QTL controlling many of the phenotypic differences between domesticated maize and teosinte, and demonstrated the existence of numerous QTL that contribute to the differences in branching architecture between the two. Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012, PAPERS}, \jri{yup, cite some PAPERS} and both \emph{tru1} and \emph{ter1} have been shown to affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tassel-replaces-upper-ears1} (\emph{tru1}), for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number) \citep{Li2012}. It seems plausible that variation in some of these other loci could have affected tillering in our greenhouse population, and contributed to the lack of correlation we see between Hopscotch genotype and tillering.

In summary, our findings demonstrate that the Hopscotch allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the Hopscotch allele is present in these populations due to recent introgression; however, it seems unlikely that  it would have drifted to high frequency in multiple populations and there may be another explanation for the high frequency we observe in some of our populations. The Hopscotch does not appear to have a strong effect reducing tillering in teosinte as it does in maize, and  other loci involved in branching architecture  may play also play roles in the regulation of tillering in teosinte. Finally, although we see no clear evidence of recent strong selection, the high frequency of the Hopscotch insertion in a number of populations continues to suggest to us that it plays an ecological role in teosinte. In the future, additional experiments will be needed to examine expression levels of \emph{tb1} and additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{ter1}) in conjunction with a more exhaustive phenotyping and genotyping assay. \jri{why not Phyb and phya? }

>>>>

%TO DO
%Fix figures and tables
%add usepackage for tables
%check citations in last paragraph for term
%reformat for journal
%finish fixing paragraph that jeff doesn't like

@@

\jri{Check citation format for AJB. I don't think it's numbers. adjust bibliography style accordingly } \lev{I would rather send to an open access journal. I don't like that no everyone would be able to have access to paper}

@@

\jri{MBH todo} Variation at \emph{tb1} has also been shown to contribute to phenotypes other than tillering \citep{Clark2006}, and a recent study by \citet{StuderDoebley2012} examined the possibility of an allelic series at the \emph{tb1} locus in teosinte. \citet{StuderDoebley2012} introgressed 9 separate teosinte segments (one from \emph{Zea diploperennis}, and four from both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis}) spanning the \emph{tb1} locus into an isogenic maize background and investigated their effects on previously associated phenotypes. They found that plants with teosinte chromosomal segments had greater tillering than their maize isogenic lines, and that different chromosomal segments of \emph{tb1} confer different amounts of tillering, suggesting that there are multiple genetic factors in this region that affect tillering. However, in addition to elucidating variance in tillering among \emph{tb1} teosinte segments, \citet{StuderDoebley2012} found significant variance among W22 control lines, suggesting that there are other genetic factors aside from alleles at the \emph{tb1} locus that affect tillering in maize. \citet{DoebleyStec1991, DoebleyStec1993} first attempted to map QTL controlling many of the phenotypic differences between domesticated maize and teosinte, and demonstrated the existence of numerous QTL that contribute to the differences in branching architecture between the two. Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012}, \jri{yup, cite some PAPERS} \lev{hah oops i missed taking out that note... :)} and both \emph{tru1} and \emph{ter1} have been shown to affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tassel-replaces-upper-ears1} (\emph{tru1}), for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number) \citep{Li2012}. It seems plausible that variation in some of these other loci could have affected tillering in our greenhouse population, and contributed to the lack of correlation we see between Hopscotch genotype and tillering.

In summary, our findings demonstrate that the Hopscotch allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the Hopscotch allele is present in these populations due to recent introgression; however, it seems unlikely that  it would have drifted to high frequency in multiple populations and there may be another explanation for the high frequency we observe in some of our populations. The Hopscotch does not appear to have a strong effect reducing tillering in teosinte as it does in maize, and  other loci involved in branching architecture  may play also play roles in the regulation of tillering in teosinte. Finally, although we see no clear evidence of recent strong selection, the high frequency of the Hopscotch insertion in a number of populations continues to suggest to us that it plays an ecological role in teosinte. In the future, additional experiments will be needed to examine expression levels of \emph{tb1} and additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{ter1}) in conjunction with a more exhaustive phenotyping and genotyping assay. \jri{why not Phyb and phya? } \lev{Are they necessary to include? I'd had them in before in a paragraph but had been voted out}
